 

### 2023-05-09 markdown/Thesis.tex
pca finished. to do umap

\begin{tabular}{l|r}
\hline
& lambda & Transformation\$\\
\hline
-2 & $\frac{1}{x^2}$
\hline
-1 & $\frac{1}{x}$
\hline
-0.5 & $\frac{1}{\sqrt{x}}$
\hline
0 & $\log{x}$
\hline
0.5 & $\sqrt{x}$
\hline
1 & $x$
\hline
2 & $x^2$
\end{tabular}

@@

The term normalization is ambiguous in data-science and will be
explained briefly in this chapter. Starting with converting the data to
the Z-distribution (Gaussian result), where 0 reflects the mean and 1
refers to the standard deviation, until vectorization where every
feature (=protein) can be explained as a vector which points in a
specific direction in the unit sphere, normalization can have different
meanings. Normalization is considered as part of the scaling procedure
and ensures that every feature contributes equally to our statistical
model and hinders large values from biasing the model in a particular
direction. Although this could also be harmful for the modeling
procedure, since the impact of important features on the dataset will
also be reduced. Depending on the algorithms later used in the data
processing, the method of choice for normalization can change the
results significantly. The K-means clustering algorithm for instance
relies on the distances between distinct data points by minimizing
variance of the squared euclidean distances. This method will be applied
in the missing value imputation which will be the next step applied to
our data. Hence we want to make sure that the get the maximum accuracy
for our imputed values. The first option to choose in the user interface
is the column-wise median and row-wise mean normalization. This method
was chosen by Specht et al in the SCoPE2 publication 2021
/citep\{Specht2021\} and worked as a reference during the development
process to benchmark other methods. When dividing each value of an
expression set by the median of the particular column, the procedure is
considered as a column-wise median normalization. Hence the
mean-normalization works the same way on the column. In the analysis
pipeline both procedures were applied on the dataset. However when the
logarithm as the transformation method was chosen in advance the
pipeline automatically switches from a division to a subtraction, since:
\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\)

Further filter steps can include correction between multiple runs. This
kind of process need the addition of a reference channel. However when
observing a single run, these steps are not crucial for the upcoming
analysis.

@@

\hypertarget{principle-component-analysis-pca}{%
\subparagraph{Principle component analysis
(PCA)}\label{principle-component-analysis-pca}}

Protein levels can be projected to their principle components (PC) and
clustered to their specific cell type. So cell types are distinguished
by their proteinaceous composition.

>>>>

\begin{center}
\begin{tabular}{l r}
$\lambda$ & Transformation \\
-2 & $\frac{1}{x^2}$ \\
-1 & $\frac{1}{x}$ \\
-0.5 & $\frac{1}{\sqrt{x}}$ \\
0 & $\log{x}$ \\
0.5 & $\sqrt{x}$ \\
1 & $x$ \\
2 & $x^2$
\end{tabular}
\end{center}

@@

The term normalization is ambiguous in data-science and will be
explained briefly in this chapter. Starting with converting the data to
the Z-distribution (Gaussian result), where 0 represents the mean and 1
represents the standard deviation, in vectorization every feature (or
protein) is represented as a vector that points in a specific direction
in the unit sphere, normalization can have different meanings.
Normalization is considered as part of the scaling procedure and ensures
that every feature contributes equally to our statistical model and
hinders large values from biasing the model in a particular direction.
However, this could also have negative consequences for the modeling
procedure, as it may reduce the impact of important features on the
dataset. Depending on the algorithms later used in the data processing,
the method of choice for normalization can change the results
significantly. The K-means clustering algorithm, for instance relies on
the distances between distinct data points by minimizing variance of the
squared euclidean distances. This method will be applied in the missing
value imputation which will be the next step applied to our data.
Therefore, we want to ensure that we obtain the most accurate imputed
values possible. The first normalization method available in the user
interface is column-wise median and row-wise mean normalization. This
method was chosen by Specht et al in the SCoPE2 publication 2021
\citep{Specht2021} and worked as a reference during the development
process to benchmark other methods. When dividing each value of an
expression set by the median of the particular column, the procedure is
considered as a column-wise median normalization. Although the
mean-normalization works the same way but takes the mean as the second
variable. In the analysis pipeline both procedures were applied on the
dataset. However when the logarithm as the transformation method was
chosen as transformation method, the pipeline automatically switches
from a division to a subtraction, since:
\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying
this normalization method, each value can be interpreted as a distance
or fraction of the mean or median of the corresponding column or row,
enabling fair comparisons between features and samples. The second
method for normalization included is the CONSTANd method, which was
proven suitable for relative quantification in the field of proteomics
\citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq
data. CONSTANd uses a technique called matrix raking, which employs the
RAS algorithm. The expression set reflects the nonnegative real
\((m, n)\) matrix \(A\) were the bi proportional constrained matrix
problem will be solved by finding the \((m, n)\) matrix \(B\) which
equals to \(diag(x) * A *diag(y)\) . Whereas \(x \in \mathbb{R}^{m}\)
and \(y \in \mathbb{R}^{n}\). The solution to this problem involves
finding the row sum of \(B=u_i\) and the column sum of \(B=v_j\), where
\(i\) and \(j\) denote the row and column indices of the matrix,
respectively \citep{Bacharach1965}. In other words, the matrix will be
alternatively manipulated on rows and columns until the mean of both
equals 1. In order to apply the method and yield true results, one has
to consider 3 major assumptions between sample types. This assumption
can be observed in the MA-plot which indicates the differences of two
samples. If we have the intensities of two samples, \(R\) and \(G\), we
can plot them on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\)
and the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
sample types of identically processed sets is needed before calculation
and the assumptions must be fulfilled.

\begin{itemize}
\item{\textbf{1}} \textbf{The majority of proteins are not differentially expressed}
\\Density of the dots decreases when looking from the middle of the cloud towards the edges.

\item{\textbf{2}} \textbf{Up- and downregulation is balanced around the mean expression}
\\The scatterplot of the data points is symmetrical around the mean on the horizontal axis.

\item{\textbf{3}} \textbf{Systematic bias correlates with the magnitude of expression}
\\The symmetry axis is approximately horizontal.

\end{itemize}

If the MA-plot does not visualize any of these violations the CONSTANd
method can be applied on the dataset and the biological meaning of the
sample types can be assessed. The method is utilized by the bioconductor
package handler. However if the dataset already consists of balanced
features and samples or the user wants to benchmark the methods applied,
normalization can be skipped also.

\hypertarget{missing-value-handling}{%
\paragraph{Missing value handling}\label{missing-value-handling}}

When analyzing multiple batches simultaneously, it is possible to
encounter situations where a particular protein is not detected in one
of the batches. From a biological perspective, it is highly unlikely
that a cell completely lacks a single protein, as it would imply a
complete loss of function for that protein. Therefore, an intensity
value of 0 is close to impossible for most proteins. In order to handle
this dark space in the expression set, several methods can be chosen in
the application. A variety of function employs replacing the missing
values with the mean or median of the rest of the matrix. This procedure
introduces no bias in the data if the missingness is low, although
differential expression for the sample carrying the missing feature can
not be expected as well. Even a more conservative approach is dropping
the rows for the feature with the missing value in one of the observed
samples. A common practice in proteomics is utilizing the K-nearest
neighbor algorithm (=KNN), which imputes the missing values by a
euclidean metrics of the neighboring values in the columns where the
feature is not missing. \(K\) denotes for the count of neighboring
values of a gene and can be selected upon preference. The euclidean
distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the
\(K\) nearest neighbors average value will be assigned to the intensity
of the missing feature. Since the average is a statistical metric which
can be biased easily by skewed distributions, normalization is an
important task before performing missing value imputation. KNN was
benchmarked for the regression of protein abundance intensities and
showed an Area under the Curve (AUC) above 0.7 when testing on a human
proteome test set \citep{Lan2013}.

\hypertarget{batch-correction}{%
\paragraph{Batch correction}\label{batch-correction}}

Since mass spectrometry experiments are highly biased by several
factors, such as sample processing, the technician, manufacturer and
production of the device (chromatography column, mass-spectrometer), the
reagents etc., working in multiple batches introduces further variance
in the protein data. This variance would overshadow the biological
differences across sample types. Before attempting to remove batch
effects it is recommended to elucidate the variance with a
dimensionality reduction technique which will be explained in the next
chapter. To remove the technical factor also called non-biological
experimental variation the R package sva \citep{Leek2012} with the
ComBat function \citep{Johnson2007} was used. The expression of a gene
\(g\) for the batch \(m_i\) and sample \(j\) is defined as
\(Y_i_j_g = a_g + X\beta_g + \gamma_i_g + \delta_i_g\epsilon_i_j_g\),
whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic
error caused by the batch. In order to remove the systematic error the
bayesian framework shrinks the effect by pooling information over the
features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
natural as they are and should reflect the true biological difference
between experimental obtained samples or groups. ComBat standardizes the
data by a Z-transformation before attempting to remove the batch effect.
By the method of moments the two batch effect parameters \(\gamma\) and
\(\delta\epsilon\) were estimated and can be subtracted from each value
for the particular feature.

@@

After cleaning the data and crystallizing from impurities such as noise
and unwanted effects, the biological differences between sample types
can be obtained. Proteomic data consists of multi dimensions (=features)
which are overall hard to understand by a glance. An approach to observe
the differences without a closer sight of exact effect on the
proteinaceous ensemble, but still keep a high level of information can
be dimensionality reduction.

\hypertarget{principle-component-analysis-pca}{%
\subparagraph{Principle component analysis
(=PCA)}\label{principle-component-analysis-pca}}

The most common approach in data science for any kind of high
dimensional data is the PCA. Every principle component can be understood
as a vector pointing in a specific direction derived as the eigenvector
from the covariance matrix of the expression set. The covariance matrix
indicates for the co linearity between the variance of all elements
compared. After calculating the covariance matrix the eigenvectors for
the covariance matrix will be obtained by solving the quadratic
equations to obtain the eigenvalues \(\lambda\). Since
\(A v = \lambda v\) the eigenvector \(v\) can be obtained by solving the
equation for each element of the vector. The eigenvectors \(v\) never
changes is direction and explains the covariance within the dataset and
so also the mathematical differences between the samples. The
eigenvectors are sorted according to their eigenvalues in decreasing
order and therefore decreasing explained variance. For graphical
visualizations the principle components are plotted against each other.
The first two components, which explain the majority of variance in the
dateset highlight the difference between samples. After the cleaning
procedures any additional effect caused by technical or biological
replication should be removed, however it could be found be highlighting
according to the particular factor. This makes the PCA to a useful tool
before starting any statistical analysis by checking for involuntarily
introduced bias.
